Exploring the conformational dynamics of a large molecule in solution using an external electric field: A nonequilibrium molecular dynamics simulation

### 2013-11-09 writedown/npt-neq/alanine.tex
revise by Luigi, mostly...

\begin{abstract}
In this paper, we investigate the conformational dynamics of alanine
   dipeptide under an external electric field by the
   nonequilibrium molecular dynamics simulation.  We consider
   the case of a constant and of an oscillatory field. In this context we
propose a procedure to
   implement the temperature control, which removes the irrelevant thermal
effects of the field. For the constant field different time-scales are
   identified  in the conformational, dipole moment, and orientational
   dynamics. Moreover, we prove that the solvent structure only
   marginally changes when the external field is switched on.  In the
   case of oscillatory field, the conformational changes are shown to be as
   strong as the previous case, and non-trivial nonequilibrium
   circular paths in the conformation space are revealed by calculating the
   integrated net probability fluxes.
  % In this paper, we investigate the conformational dynamics of alanine
  % dipeptide under an external electric field (EF) by the
  % nonequilibrium molecular dynamics (MD) simulation, which enables us
  % to calculate the nonequilibrium physical properties that are not
  % available for traditional equilibrium MD simulation.  We study both
  % the constant and oscillatory EF cases, and discuss the way to
  % implement the temperature control in the nonequilibrium MD
  % simulation, which removes the irrelevant thermal effect of the EF
  % from the system. In the constant EF case, different time-scales are
  % discovered in the conformational, dipole moment, and orientational
  % dynamics. Moreover, we prove that the solvent structure only
  % marginally changes when the external EF is switched on.  In the
  % oscillatory EF case, the conformational changes are shown to be as
  % strong as constant EF case, and non-trivial nonequilibrium
  % conformational circulations are presented by calculating the
  % integrated net probability fluxes.
\end{abstract}

@@

The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental e.g.~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical e.g.~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects}
studies have been devoted to this topic.

In this perspective, molecular dynamics (MD) has been proven useful in
understanding the behavior of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum
time range of the order of 10-60 nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), the dipole
moment, shape parameters are calculated as a function of time, or
in stationary condition, taking time averages.
This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
the behavior of observables in time is characterized by 
strong
fluctuations. The entity of the fluctuations is such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
A time average can reduce the statistical uncertainty,
but in non-stationary conditions, it is a not-well-founded procedure.
In fact, we are studying non-stationary nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).
In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it.
This is the approach that will be followed in this work, using
the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method, firstly developed by one of us and coworkers~\cite{ciccotti1975direct, ciccotti1979thought, ciccotti1993theoretical, palla2008bulk}, 
has been recently applied, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is to
provide a way of evaluating time-dependent
nonequilibrium averages starting from a properly determined initial ensemble.
In this context, the method can be used to analyze the
conformational changes of a solvated molecule, subject to a time-dependent EF.
In the present paper, we study a small peptide, alanine dipeptide, as a model for showing how the method can be used. 
The choice of a small peptide, rather than a larger one, is done firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding the slow-going interplay between
different peptide segments along a chain.
Secondly, the computational cost required is rather massive and thus  testing
the various methodology proposed here would be hampered for large molecules,
although they \bluec{remain (\textit{does Giovanni write ``remain''? I cannot read})} our goal.

The paper is organized as follows:
\recheck{
  We firstly describe the idea of D-NEMD, and a local thermostating method
for nonequilibrium temperature control.
% Then
% introduce the physical
% observables chosen to analyze the nonequilibrium responses.
Then
after introducing the physical
observables chosen to analyze the nonequilibrium response,
the numerical results for the response of alanine
dipeptide under static and oscillatory EF are
finally presented.} By calculating the nonequilibrium properties from D-NEMD
simulations, we are able to address
a few relavant questions: (i)~How the conformation of the peptide
changes in an EF (conformational space);
(ii)~When do these changes happen (time scale),
and how likely they are (probability). 
To our knowledge, this study is one of the few (if not the first)  where
time-dependent nonequilibrium w.r.t. conformation
is explicitly considered for a \bluec{large molecule \textit{what does Giovanni write here? For me, the alanine dipeptide molecule is NOT large at all!}}
in solution subject to time-dependent EF.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the dynamical approach to nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known. In particular, in our case,
it is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that he dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x, t)$, where $\mathcal H$ is the time-dependent Hamiltonian,
and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = U^\dagger(t,0) \rho(\vect x, 0)$,
where $U^\dagger(t,0) = \exp\{-i \int_0^t d t' L(t')\}$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by
$\hat O(\vect x(t)) = U(t,0)\,\hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), U^\dagger(t,0)\, \rho(\vect x, 0)\rangle
  = \langle U(t,0)\,\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the distribution of equilibrium (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable computed along trajectories, starting from
initial configurations sampled from an \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the full Hamiltonian dynamics is integrated until
time $t$.
In this paper, these trajectories are called \emph{branching
  trajectories} (see Fig.\ref{branching} for a pictorial representation). 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (e.g.~Langevin dynamics) can be handled analogoursly.

%Being more generalized, if the microscopic observable depends not only on the end
%configuration at time $t$, but also on the historical
%configurations along the trajectory, then expression~\eqref{eqn:tmp4} takes the generalized form:
%\begin{align}
%  O(t) = \int d\vect x\,\rho(\vect x, 0) \int_{\mathcal C\{\vect x, 0; t\}} \hat O[\vect x_s] \,d\mathcal P[\vect x_s] 
%\end{align}
%where the microscopic observable $ \hat O[\vect x_s] $ is now a
%functional of the trajectory.  $\mathcal C\{\vect x, 0; t\}$ is the
%set of all continuous trajectories starting at point $\vect x$ at time
%0, and ending at time $t$. $\mathcal P[\vect x_s] $ is the measure of
%the trajectory space $\mathcal C$.  In our case, it is the
%$\delta$-function peaked around the trajectories governed by the
%Hamiltonian dynamics. To be general, it can also be the measure of all
%trajectories generated by, for example, a Langevin dynamics, then the
%branching trajectories should be generated by the Langevin
%dynamics rather than the Hamiltonian dynamics.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control scheme.
    The white area is the region characterized by pure dynamics without any direct action by the external thermostat.
    The gray area is the region where the thermostat is used to keep the target temperature.
    $R_{ex}$ denotes the size of the purely dynamical region, which
    is a sphere around the molecule of interest (in our case it is the alanine
    dipeptide).
    % at the center of the box.
    The whole system is subject to periodic boundary conditions. }
  \label{fig:tmp2}
\end{figure}

% The aforementioned D-NEMD simulation algorithm
% does not describe how the boundary of the system should be set up.
% The very naive idea of doing the simulation
% is to simulate a infinitely large isolated system, or any large enough
% isolated subsystem of it.
% In practice, 
% Therefore, we are interested in a 
For this study, we are interested in 
nonthermal~\cite{delaHoz2005microwaves} effects of the EF
on the configuration of the solvated alanine dipeptide.
For this reason a proper control of the temperature in a nonequilibrium system
becomes an important issue; in fact it was argued that  an
accurate control of the internal reaction temperature is essential to
perform reproducible experiments~\cite{damm2012can}.
A straightforward (and realistic) implementation of the temperature control consists in 
embedding the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine. In this case
any extra heat generated by the EF can be effectively
dispersed in the large solvent environment.
In this way one can identify the effect of the EF on the conformational space of the solvated molecule,
without the artifact that the surrounding solvent cannot release the additional energy of the EF.
However, in practice, in most of the cases the computational cost of having large solvent environment is rather massive if not prohibitive.
Therefore, it is of interest to substitute the infinitely large
solvent environment with a smaller affordable system and at the same time accurately reproduce the nonequilibrium
physical properties of interest.
In MD, for equilibrium situations, the solution is to couple
the system to an external thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by simulating a finite size system with periodic boundary
conditions.
However, coupling the system to a thermostat for nonequilibrium situations is a more delicate issue.
In fact, since a thermostat is explicitly designed to bring the system into
thermal equilibrium,
the perturbation produced by the EF would be mostly adsorbed by the thermostat. This means that the physics of the original dynamics in presence of the EF
would be lost and the response to the perturbation would be unphysical. 
%will disturb the original Hamiltonian dynamics. This is
%in general problematic, 
%because the dynamics  under thermostating is artificial designed,
%so that the equilibrium distribution can be correctly produce, but not
%the nonequilibrium properties of system.
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% However, for a nonequilibrium system, the \emph{dynamics} properties
% will be distrubed by the thermostat, because the
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% For a nonequilibrium simulation, simply couple the system
% to a thermostat may not be a good idea, because the \emph{dynamics}
% generated by the thermostat is artificial, and deviation from the
% Hamiltonian dynamics under investigation will introduce artificial
% a effects in the nonequilibrium observations.

In this section we propose a procedure to solve the problem: we
start observing that the dynamics without thermostat of the alanine, 
and the water molecules in the closer solvation shells, is
crucial to determine the response of the observable of interest
under the EF perturbation
(in this work, the observable is the conformational change of alanine).
Instead the detailed dynamics of
the water molecules far away is not of primary importance: the region beyond the first solvation shells plays mainly the role of a thermodynamic bath. 
Therefore, instead of simulating an infinitely
large system, we simulate our finite size system with periodic
boundary condition, and with the following characteristics:
(i)~The simulating region is divided in two
subregions, a spherical region around the alanine dipeptide of radius $R_{ex}$,
centered at the alpha-carbon. Here the dynamics is
not subject to a thermostat.
We refer to such subregion as \emph{the  dynamical
region}. (ii)~Beyond $R_{ex}$, the dynamics of the water molecules is coupled
to a Langevin thermostat.  This region is called \emph{the thermostated
region}. For technical convenience we
fix the position of the alpha-carbon in space, so that the alanine is always
located at the center of the simulation region (see Fig.~\ref{fig:tmp2}).
In these conditions, the  dynamics is preserved in the dynamical region (where the properties of interest are observed), while 
possible artificial effects of the thermostating process in the outside region
are negligible due to the finite correlation
range of liquid water.
At the same time, the thermostated region works as a infinitely large
environment that effectively absorbs the extra heat produced by the EF.
The validity of the above
statement will be checked by a series of numerical tests where it is shown that
the response of observables in nonequilibrium (i.e. under the effect of the EF) does not depend on the size of the system
and on the size of the  dynamical subregion, provided that they are
reasonably large.

% Notice again we actually
% do not want to precisely calculate each trajectory starting from the
% initial configurations,
% but want to correctly calculate the nonequilibrium observable,
% we

@@

\subsection{System settings and simulation protocol}
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and dissolved in 644 TIP3P
water molecules.
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  dynamical region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
First, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
% They server as starting point of the branching
% trajectories.
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by Gromacs.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat~\cite{parrinello1981polymorphic} with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizeable effect on the
dynamics of the system.
At time $t=0$~ps, the system
has been fully equilibrated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the EF is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, for the case of constant EF, we have
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ for $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
for $t \geq t_{warm}$.

%  (see also Fig.~\ref{fig:tmp3} for
%  the vanished initial dipole moment)

@@

\begin{figure}
  \centering
  \includegraphics[]{fig-meta-npt.eps}
  \caption{\recheck{The time-dependent probability of conformations $P_I(t)$} 
    when a constant EF is present.
    The warm-up time is $t_{warm} = 10$~ps. The red line refers to conformation $I = A_1$,
    green to $A_2$, dark blue to $B_1$, purple to $B_2$ and light blue
    to $C$. \recheck{Please notice that the summation of all conformational
      probabilities
    is identity, i.e.~$\sum_I P_I(t) = 1$.}
  }
  \label{fig:tmp5}
\end{figure}

>>>>

\begin{abstract}
In this paper, we investigate the conformational dynamics of alanine
   dipeptide under an external electric field by 
   nonequilibrium molecular dynamics simulation.  We consider
   the case of a constant and of an oscillatory field. In this context we
propose a procedure to
   implement the temperature control, which removes the irrelevant thermal
effects of the field. For the constant field different time-scales are
   identified  in the conformational, dipole moment, and orientational
   dynamics. Moreover, we prove that the solvent structure only
   marginally changes when the external field is switched on.  In the
   case of oscillatory field, the conformational changes are shown to be as
   strong as the previous case, and non-trivial nonequilibrium
   circular paths in the conformation space are revealed by calculating the
   integrated net probability fluxes.

\end{abstract}

@@

The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental e.g.~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical e.g.~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects}
studies have been devoted to this topic.
In this perspective, molecular dynamics (MD) has been proven useful in
understanding the behaviour of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum
time range of the order of 10-60 nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), dipole
moment, shape parameters are calculated as a function of time, or
in stationary condition, taking time averages.
This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
the behaviour of observables in time is characterized by 
strong
fluctuations. The entity of the fluctuations is such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
A time average can reduce the statistical uncertainty,
however in non-stationary conditions this would be a not-well-founded procedure.
In fact, we are studying non-stationary nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).
In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it.
This is the approach that will be followed in this work, using
the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method, firstly developed by one of us and coworkers~\cite{ciccotti1975direct, ciccotti1979thought, ciccotti1993theoretical, palla2008bulk}, 
has been recently applied, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is to
provide a way of evaluating time-dependent
nonequilibrium averages starting from a properly determined initial ensemble.
In this context, the method can be used to analyze the
conformational changes of a solvated molecule, subject to a time-dependent EF.
In the present paper, we study a small peptide, alanine dipeptide, as a model for showing how the method can be used. 
The choice of a small peptide, rather than a larger one, is done firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding the slow-going interplay between
different peptide segments along a chain. 
Secondly, the computational cost required is rather massive and thus  testing
the various methodology proposed here would be hampered for large molecules,
although they remain our goal.  \recheck{In any case, alanine dipeptide
is large enough so that the conformational changes of its internal degrees of freedom represent valid observables for a test of validity and utility of our approach.}
The paper is organized as follows:
\recheck{
  We firstly describe the idea of D-NEMD, and propose a technical improvement consisting in the definition and application of a local thermostating process.
Next, after introducing the physical
observables chosen to analyze the nonequilibrium response,
the numerical results for the case of a static and oscillatory EF, respectively, are discussed.} By calculating the nonequilibrium properties from D-NEMD
simulations, we are able to address
a few relevant questions: (i)~How the conformation of the peptide
changes in an EF (conformational space);
(ii)~When these changes happen (time scale),
and how likely they are (probability). 
To our knowledge, this study is one of the few (if not the first)  where
time-dependent nonequilibrium w.r.t. conformation
is explicitly considered for a (relatively large) molecule
in solution subject to an external time-dependent EF.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the dynamical approach to nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known. In particular, in our case,
it is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that he dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x, t)$, where $\mathcal H$ is the time-dependent Hamiltonian,
and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = U^\dagger(t,0) \rho(\vect x, 0)$,
where $U^\dagger(t,0) = \exp\{-i \int_0^t d t' L(t')\}$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by
$\hat O(\vect x(t)) = U(t,0)\,\hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), U^\dagger(t,0)\, \rho(\vect x, 0)\rangle
  = \langle U(t,0)\,\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the distribution of equilibrium (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable computed along trajectories, starting from
initial configurations sampled from an \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the full Hamiltonian dynamics is integrated until
time $t$.
In this paper, these trajectories are called \emph{branching
  trajectories} {\bf EITHER WE PUT THE PICTURE OR REMOVE THIS SENTENCE {(see Fig.\ref{branching} for a pictorial representation)}}. 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (e.g.~Langevin dynamics) can be handled analogously.

@@

  \includegraphics[width=0.3\textwidth]{thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control scheme.
    The white area is the region characterized by pure dynamics without any direct action by the external thermostat.
    The gray area is the region where the thermostat is used to keep the target temperature.
    $R_{ex}$ denotes the size of the purely dynamical region, which
    is a sphere around the molecule of interest (in our case it is the alanine
    dipeptide).
    % at the center of the box.
    The whole system is subject to periodic boundary conditions. }
  \label{fig:tmp2}
For this study, we are interested in 
nonthermal~\cite{delaHoz2005microwaves} effects of the EF
on the configuration of the solvated alanine dipeptide.
For this reason a proper control of the temperature in a nonequilibrium system
becomes an important issue; in fact it was argued that  an
accurate control of the internal reaction temperature is essential to
perform reproducible experiments~\cite{damm2012can}.
A straightforward (and realistic) implementation of the temperature control consists in 
embedding the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine. In this case
any extra heat generated by the EF can be effectively
dispersed in the large solvent environment.
In this way one can identify the effect of the EF on the conformational space of the solvated molecule,
without the artifact that the surrounding solvent cannot release the additional energy of the EF.
However, in practice, in most of the cases the computational cost of having large solvent environment is rather massive if not prohibitive.
Therefore, it is of interest to substitute the infinitely large
solvent environment with a smaller affordable system and at the same time accurately reproduce the nonequilibrium
physical properties of interest.
In MD, for equilibrium situations, the solution is to couple
the system to an external thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by simulating a finite size system with periodic boundary
conditions.
However, coupling the system to a thermostat for nonequilibrium situations is a more delicate issue.
In fact, since a thermostat is explicitly designed to bring the system into
thermal equilibrium,
the perturbation produced by the EF would be mostly adsorbed by the thermostat. This means that the physics of the original dynamics in presence of the EF
would be lost and the response to the perturbation would be unphysical. 
In this section we propose a procedure to solve the problem: we
start observing that the dynamics without thermostat of the alanine, 
and the water molecules in the closer solvation shells, is
crucial to determine the response of the observable of interest
under the EF perturbation
(in this work, the observable is the conformational change of alanine).
Instead the detailed dynamics of
the water molecules far away is not of primary importance: the region beyond the first solvation shells plays mainly the role of a thermodynamic bath. 
Therefore, instead of simulating an infinitely
large system, we simulate our finite size system with periodic
boundary condition, and with the following characteristics:
(i)~The simulating region is divided in two
subregions, a spherical region around the alanine dipeptide of radius $R_{ex}$,
centered at the alpha-carbon. Here the dynamics is
not subject to a thermostat.
We refer to such subregion as \emph{the  dynamical
region}. (ii)~Beyond $R_{ex}$, the dynamics of the water molecules is coupled
to a Langevin thermostat.  This region is called \emph{the thermostated
region}. For technical convenience we
fix the position of the alpha-carbon in space, so that the alanine is always
located at the center of the simulation region (see Fig.~\ref{fig:tmp2}).
In these conditions, the  dynamics is preserved in the dynamical region (where the properties of interest are observed), while 
possible artificial effects of the thermostating process in the outside region
are negligible due to the finite correlation
range of liquid water.
At the same time, the thermostated region works as a infinitely large
environment that effectively absorbs the extra heat produced by the EF.
The validity of the above
statement will be checked by a series of numerical tests where it is shown that
the response of observables in nonequilibrium (i.e. under the effect of the EF) does not depend on the size of the system
and on the size of the  dynamical subregion, provided that they are
reasonably large.

@@

\subsection{System settings and simulation protocol}
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and dissolved in 644 TIP3P
water molecules.
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  dynamical region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
First, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by GROMACS.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat~\cite{parrinello1981polymorphic} with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizable effect on the
dynamics of the system.
At time $t=0$~ps, the system
has been fully equilibrated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the EF is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, for the case of constant EF, we have
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ for $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
for $t \geq t_{warm}$.

@@

\begin{figure}
  \centering
  \includegraphics[]{fig-meta-npt.eps}
  \caption{\recheck{The time-dependent probability of conformations $P_I(t)$ 
    when a constant EF is present, where $\sum_I P_I(t) = 1, \forall t$.}
    The warm-up time is $t_{warm} = 10$~ps. The red line refers to conformation $I = A_1$,
    green to $A_2$, dark blue to $B_1$, purple to $B_2$ and light blue
    to $C$. 
  }
  \label{fig:tmp5}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig-trans-010-flux-1.eps}
  \includegraphics[width=0.3\textwidth]{fig-trans-010-flux-2.eps}
  \includegraphics[width=0.3\textwidth]{fig-trans-010-flux-3.eps}
  \includegraphics[width=0.3\textwidth]{fig-trans-010-flux-4.eps}
  \includegraphics[width=0.3\textwidth]{fig-trans-010-flux-5.eps}
  \caption{\recheck{
    The net probability flux for the constant EF case.
    The warm-up time is $t_{warm} = 10$~ps.
    The probability flux $F_{J,I}(t)$ (defined by Eqn.~\eqref{eqn:tmp7}) is reported in units of $\textrm{ps}^{-1}$.
    From top, left to right we report $I = A_1$, $A_2$, $B_1$; bottom, left to right $B_2$ and
    $C$, respectively. In each plot, the red line stands for $J=A_1$,
    green for $J=A_2$, dark blue for $J=B_1$, purple for $J=B_2$ and light blue
    for $J=C$.} 
    }
  \label{fig:tmp6}
\end{figure}
